### 2019-09-25 reports/thesis/chapters/3_compoundPoisson.tex
Check algebra
Revise

Returning back to the compound Poisson distribution with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$, let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting $\theta$ to zero. In other words, $\expectation[X^r]=M_X^{(r)}(0)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
  \\
  \expectation\left[(X-\expectation[X])^3\right] &=\lambda\expectation\left[U^3\right]
\end{align}
which agrees with the expectation and variance results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realisations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then was normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

@@

\begin{figure}
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonNormlambda1alpha1beta1.eps}
        \caption{Histogram - Normal approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonNormlambda1alpha1beta1.eps}
        \caption{Q-Q plot - Normal approx.}
    \end{subfigure}
    }
    \vspace{1em}
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonSaddlelambda1alpha1beta1.eps}
        \caption{Histogram - Saddlepoint approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonSaddlelambda1alpha1beta1.eps}
        \caption{Q-Q plot - Saddlepoint approx.}
    \end{subfigure}
    }
    \caption{1\,000 $\CPoisson(1,1,1)$ random variables were simulated and its empirical density is compared to the approximate p.d.f. On the left, the dotted red line shows the 68\% confidence interval of the expected frequency density.}
    \label{fig:compoundPoisson_histogram_approx_1_1_1}
\end{figure}

\begin{figure}
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonNormlambda1alpha100beta1.eps}
        \caption{Histogram - Normal approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonNormlambda1alpha100beta1.eps}
        \caption{Q-Q plot - Normal approx.}
    \end{subfigure}
    }
    \vspace{1em}
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonSaddlelambda1alpha100beta1.eps}
        \caption{Histogram - Saddlepoint approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonSaddlelambda1alpha100beta1.eps}
        \caption{Q-Q plot - Saddlepoint approx.}
    \end{subfigure}
    }
    \caption{1\,000 $\CPoisson(1,100,1)$ random variables were simulated and its empirical density is compared to the approximate p.d.f. On the left, the dotted red line shows the 68\% confidence interval of the expected frequency density.}
    \label{fig:compoundPoisson_histogram_approx_1_100_1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{../figures/compoundpoisson/graphicalModel.pdf}
  \caption{Graphical model of the grey value $X_{i,j}$ for each of the $m$ pixels in the $n$ replicate projections. $Y_{i,j}\sim\poisson(\lambda_i)$ is the photon count. The grey value has a compound Poisson gamma element $X_{i,j}|Y_{i,j}\sim\gammaDist(Y_{i,j}\alpha,\beta)$ and can be extended by adding electronic noise $\epsilon_{i,j}\sim\normal(a,\kappa)$.}
  \label{fig:compoundPoisson_graphicalModel}
\end{figure}

@@

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
  \label{eq:compoundPoisson_jointLogLikelihood}
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions were met.

@@

In the E step, the realisations of $Y$ are estimated using
\begin{equation}
  y_i =
  \expectation\left[
    Y|X=x_i
  \right]
\end{equation}
given the parameters $\lambda$, $\alpha$ and $\beta$. The conditional expectation is calculated using
\begin{equation}
  y_i = 
  \begin{cases}
    0 & \text{ for } x=0 \\ 
    \dfrac{\sum_{y=1}^\infty y \prob(Y=y|X=x_i)}{\sum_{y=1}^\infty \prob(Y=y|X=x_i)} & \text{ for } x>0
  \end{cases}
\end{equation}
where
\begin{equation*}
  \prob(Y=y|X=x) = \frac{p_{X|Y}(x|y)\prob(Y=y)}{p_X(x)}
  \ .
\end{equation*}
Focussing on the $x>0$ case for now
\begin{equation*}
  \prob(Y=y|X=x) = \frac{1}{p_X(x)}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\frac{\euler^{-\lambda}\lambda^y}{y!}
\end{equation*}
which is simplified to
\begin{equation}
  \prob(Y=y|X=x) = W_y \frac{\euler^{-\lambda-\beta x}}{x p_X(x)} \ .
\end{equation}
Then, the conditional expectation is
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty y W_y}{\sum_{y=1}^\infty W_y}
  \ .
\end{equation}
As discussed before, the sum in the denominator can be evaluated using the method by \cite{dunn2005series}. A similar method for evaluating the numerator can be obtained by truncating the sum and summing over large terms.

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\dotsc
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly,
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated in the M step so that they can be used in the E step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. $\ln(W^{(r)}_y)$ can be expressed as
\begin{equation}
  \ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and taking the derivative with respect to $y$ obtains
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotsc$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y}
  \ .
\end{equation}
Therefore, the maximum of $W_y^{(r)}$ is located at $y_\text{max}=\dfrac{x^{2-p}}{\phi(2-p)}$ for all $r=0,1,2,\dotsc$. As a result, the same method for evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\dotsc$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@

The conditional expectation $y_i = \expectation[Y_i|X_i=x_i]$ and $\zeta_i = \variance[Y_i|X_i=x_i]$ were calculated beforehand in the E step. The quantity $\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]$ can be calculated using the approximation
\begin{equation}
  \expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i] \approx
  \ln\Gamma(\alpha y_i) + \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
\end{equation}
where $\psi(n)$ is the digamma function. The objective function is then
\begin{multline}
  T(\lambda,\alpha,\beta)=
  -n\lambda
  +\sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\alpha\ln\beta-\ln\Gamma(\alpha y_i) - \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \beta x_i
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
  \ .
\end{multline}

@@

Maximising $T$ with respect to $\alpha$ and $\beta$ can be done numerically using the Newton-Raphson method since derivatives up to the second-order can be obtained. For the first-order derivatives, these are:
\begin{multline}
  \frac{\partial T}{\partial \alpha} =
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    y_i\ln\beta -\psi(\alpha y_i)y_i-\zeta_i\alpha\psi'(\alpha y_i)
    \vphantom{\frac{1}{1}}
  \right.
  \\
  \left.
    - \frac{1}{2}\zeta_i\alpha^2\psi''(\alpha y_i)y_i + y_i\ln x_i
  \right]
\end{multline}
and
\begin{equation}
  \frac{\partial T}{\partial \beta} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
  \frac{\alpha y_i}{\beta}-x_i
  \right]
  \ .
\end{equation}
The second-order derivatives are:
\begin{equation}
  \frac{\partial^2 T}{\partial \alpha \partial \beta} =
  \sum_{i=1}^n \mathbb{I}(x_i>0)\left[\frac{y_i}{\beta}\right]
  \ ,
  \label{eq:compoundPoisson:d2tdadb}
\end{equation}
\begin{equation}
  \frac{\partial^2 T}{\partial \beta^2} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[-\frac{\alpha y_i}{\beta^2}
  \right]
  \label{eq:compoundPoisson:d2td2b}
\end{equation}
and
\begin{multline*}
  \frac{\partial^2 T}{\partial \alpha^2} = 
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    -y_i^2\psi'(\alpha y_i) - \zeta_i\psi'(\alpha y_i) - \zeta_i\alpha y_i\psi''(\alpha y_i)
    \vphantom{\frac{1}{2}}\right.\\\left. 
    - \zeta_i\alpha\psi''(\alpha y_i)y_i
    -\frac{1}{2}\zeta_i\alpha^2\psi'''(\alpha y_i)y_i^2
  \right]
\end{multline*}
simplifying to
\begin{multline}
  \frac{\partial^2 T}{\partial \alpha^2} = 
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    -(y_i^2+\zeta_i)\psi'(\alpha y_i) - 2\zeta_i\alpha y_i\psi''(\alpha y_i)
    \vphantom{\frac{1}{2}}
  \right.
  \\
  \left.  
    -\frac{1}{2}\zeta_i\alpha^2y_i^2\psi'''(\alpha y_i)
  \right] \ .
  \label{eq:compoundPoisson:d2td2a}
\end{multline}

@@

For $\lambda=10$ and $\lambda=100$ case, the estimates of $\alpha$ and $\beta$ struggled to converge and increased/decreased without bounds without affecting the log-likelihood. It appeared that the EM algorithm failed for $\lambda>10$ looking at these particular examples.

\begin{figure}[p]
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_lnL.eps}
        \caption{Log-likelihood}
    \end{subfigure}
        \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_lambda.eps}
        \caption{$\lambda$}
    \end{subfigure}
    }
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_alpha.eps}
        \caption{$\alpha$}
    \end{subfigure}
        \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_beta.eps}
        \caption{$\beta$}
    \end{subfigure}
    }
    \caption{EM algorithm was used to estimate the parameters of a $\CPoisson(1,1,1)$ random variable using 1\,000 simulated samples, repeated 10 times. The graphs show the log-likelihood and the estimated parameters at each EM step for each repeat of the experiment. The dashed lines show the standard deviation using the Cram\'er-Rao lower bound.}
    \label{fig:compoundPoisson_convergence_1}
\end{figure}

>>>>

Returning to the compound Poisson distribution with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$, let the moment generating function (m.g.f.) of $X$ be $M_X(\theta)=\expectation\left[\euler^{X\theta}\right]$. It can be shown that the m.g.f.~is
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
\end{equation}
\citep{gatto2010saddlepoint}. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}. Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting $\theta$ to zero. In other words, $\expectation[X^r]=M_X^{(r)}(0)$. Then it can be shown that
\begin{align}
  \expectation\left[X\right]&=\lambda\expectation\left[U\right]
  \\
  \variance\left[X\right] &= \lambda\expectation\left[U^2\right]
  \\
  \expectation\left[(X-\expectation[X])^3\right] &=\lambda\expectation\left[U^3\right]
\end{align}
which agrees with the expectation and variance results in the previous section. The derivation is shown in Appendix \ref{chapter:appendix_compoundPoissonMgf}.

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realisations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

@@

\begin{figure}[t]
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonNormlambda1alpha1beta1.eps}
        \caption{Histogram - Normal approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonNormlambda1alpha1beta1.eps}
        \caption{Q-Q plot - Normal approx.}
    \end{subfigure}
    }
    \vspace{1em}
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonSaddlelambda1alpha1beta1.eps}
        \caption{Histogram - Saddlepoint approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonSaddlelambda1alpha1beta1.eps}
        \caption{Q-Q plot - Saddlepoint approx.}
    \end{subfigure}
    }
    \caption{1\,000 $\CPoisson(1,1,1)$ random variables were simulated and its empirical density is compared to the approximate p.d.f. On the left, the dotted red line shows the 68\% confidence interval of the expected frequency density.}
    \label{fig:compoundPoisson_histogram_approx_1_1_1}
\end{figure}

\begin{figure}[t]
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonNormlambda1alpha100beta1.eps}
        \caption{Histogram - Normal approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonNormlambda1alpha100beta1.eps}
        \caption{Q-Q plot - Normal approx.}
    \end{subfigure}
    }
    \vspace{1em}
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_CompoundPoissonSaddlelambda1alpha100beta1.eps}
        \caption{Histogram - Saddlepoint approx.}
    \end{subfigure}
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/cpHistogram_qq_CompoundPoissonSaddlelambda1alpha100beta1.eps}
        \caption{Q-Q plot - Saddlepoint approx.}
    \end{subfigure}
    }
    \caption{1\,000 $\CPoisson(1,100,1)$ random variables were simulated and its empirical density is compared to the approximate p.d.f. On the left, the dotted red line shows the 68\% confidence interval of the expected frequency density.}
    \label{fig:compoundPoisson_histogram_approx_1_100_1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{../figures/compoundpoisson/graphicalModel.pdf}
  \caption{Graphical model of the grey value $X_{i,j}$ for each of the $N$ pixels in the $m$ replicate projections. $Y_{i,j}\sim\poisson(\lambda_i)$ is the photon count. The grey value has a compound Poisson gamma element $X_{i,j}|Y_{i,j}\sim\gammaDist(Y_{i,j}\alpha,\beta)$ and can be extended by adding electronic noise $\epsilon_{i,j}\sim\normal(a,\kappa)$.}
  \label{fig:compoundPoisson_graphicalModel}
\end{figure}

@@

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
  \label{eq:compoundPoisson_jointLogLikelihood}
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$'s given the $X$'s and parameters (E step), followed by estimating the parameters given the $Y$'s and $X$'s (M step) until some convergence conditions were met.

@@

In the E step, the realisations of $Y$ are estimated using
\begin{equation}
  y_i =
  \expectation\left[
    Y|X=x_i
  \right]
\end{equation}
given the parameters $\lambda$, $\alpha$ and $\beta$. The conditional expectation is calculated using
\begin{equation}
  y_i = 
  \begin{cases}
    0 & \text{ for } x=0 \\ 
    \dfrac{\sum_{y=1}^\infty y \prob(Y=y|X=x_i)}{\sum_{y=1}^\infty \prob(Y=y|X=x_i)} & \text{ for } x>0
  \end{cases}
\end{equation}
where
\begin{equation*}
  \prob(Y=y|X=x) = \frac{p_{X|Y}(x|y)\prob(Y=y)}{p_X(x)}
  \ .
\end{equation*}
Focussing on the $x>0$ case for now
\begin{equation*}
  \prob(Y=y|X=x) = \frac{1}{p_X(x)}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\frac{\euler^{-\lambda}\lambda^y}{y!}
\end{equation*}
which is simplified to
\begin{equation}
  \prob(Y=y|X=x) = W_y \frac{\euler^{-\lambda-\beta x}}{x p_X(x)} \ .
\end{equation}
As a result, the conditional expectation is
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty y W_y}{\sum_{y=1}^\infty W_y}
  \ .
\end{equation}
As discussed before, the sum in the denominator can be evaluated using the method by \cite{dunn2005series}. A similar method for evaluating the numerator can be obtained by truncating the sum and summing over large terms.

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\dotsc
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly,
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated here so that they can be used in the M step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. $\ln\left(W^{(r)}_y\right)$ can be expressed as
\begin{equation}
  \ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and taking the derivative with respect to $y$ obtains
\begin{equation}
  \frac{\partial }{\partial y} \ln\left(W_y^{(r)}\right) = \frac{r}{y } + \frac{\partial }{\partial y}\ln\left(W_y\right)
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotsc$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial }{\partial y} \ln\left(W_y^{(r)}\right)\approx \frac{\partial }{\partial y} \ln\left(W_y\right)
  \ .
\end{equation}
Therefore, the maximum of $W_y^{(r)}$ is located at $y_\text{max}=\dfrac{x^{2-p}}{\phi(2-p)}$ for all $r=0,1,2,\dotsc$. As a result, the same method for evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\dotsc$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@

The conditional expectation $y_i = \expectation[Y_i|X_i=x_i]$ and $\zeta_i = \variance[Y_i|X_i=x_i]$ were calculated beforehand in the E step. The quantity $\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]$ can be calculated using the approximation
\begin{equation}
  \expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i] \approx
  \ln\Gamma(\alpha y_i) + \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
\end{equation}
where $\psi(n)$ is the digamma function. The objective function is then
\begin{multline}
  T(\lambda,\alpha,\beta)\approx
  -n\lambda
  +\sum_{i=1}^n
  \mathbb{I}(x_i>0)
  \left[
    y_i\alpha\ln\beta-\ln\Gamma(\alpha y_i) - \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
  \right.
  \\
  \left.
    +y_i\alpha\ln x_i - \beta x_i
    + y_i \ln \lambda
    \vphantom{\frac{1}{1}}
  \right]
  + c
  \ .
\end{multline}

@@

Maximising $T$ with respect to $\alpha$ and $\beta$ can be done numerically using the Newton-Raphson method since derivatives up to the second-order can be obtained. For the first-order derivatives, these are:
\begin{multline}
  \frac{\partial T}{\partial \alpha} =
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    y_i\ln\beta -\psi(\alpha y_i)y_i-\zeta_i\alpha\psi'(\alpha y_i)
    \vphantom{\frac{1}{1}}
  \right.
  \\
  \left.
    - \frac{1}{2}\zeta_i\alpha^2\psi''(\alpha y_i)y_i + y_i\ln x_i
  \right]
\end{multline}
and
\begin{equation}
  \frac{\partial T}{\partial \beta} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
  \frac{\alpha y_i}{\beta}-x_i
  \right]
  \ .
\end{equation}
The second-order derivatives are
\begin{equation}
  \frac{\partial^2 T}{\partial \alpha \partial \beta} =
  \sum_{i=1}^n \mathbb{I}(x_i>0)\left[\frac{y_i}{\beta}\right]
  \ ,
  \label{eq:compoundPoisson:d2tdadb}
\end{equation}
\begin{equation}
  \frac{\partial^2 T}{\partial \beta^2} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[-\frac{\alpha y_i}{\beta^2}
  \right]
  \ ,
  \label{eq:compoundPoisson:d2td2b}
\end{equation}
and
\begin{multline*}
  \frac{\partial^2 T}{\partial \alpha^2} = 
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    -y_i^2\psi'(\alpha y_i) - \zeta_i\psi'(\alpha y_i) - \zeta_i\alpha y_i\psi''(\alpha y_i)
    \vphantom{\frac{1}{2}}\right.\\\left. 
    - \zeta_i\alpha\psi''(\alpha y_i)y_i
    -\frac{1}{2}\zeta_i\alpha^2\psi'''(\alpha y_i)y_i^2
  \right]
\end{multline*}
simplifying to
\begin{multline}
  \frac{\partial^2 T}{\partial \alpha^2} = 
  \sum_{i=1}^n\mathbb{I}(x_i>0)
  \left[
    -(y_i^2+\zeta_i)\psi'(\alpha y_i) - 2\zeta_i\alpha y_i\psi''(\alpha y_i)
    \vphantom{\frac{1}{2}}
  \right.
  \\
  \left.  
    -\frac{1}{2}\zeta_i\alpha^2y_i^2\psi'''(\alpha y_i)
  \right] \ .
  \label{eq:compoundPoisson:d2td2a}
\end{multline}

@@

For $\lambda=10$ and $\lambda=100$, the estimates of $\alpha$ and $\beta$ struggled to converge and increased/decreased without bounds without affecting the log-likelihood. It appeared that the EM algorithm failed for $\lambda>10$ looking at these particular examples.

\begin{figure}[p]
  \centering
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_lnL.eps}
        \caption{Log-likelihood}
    \end{subfigure}
        \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_lambda.eps}
        \caption{$\lambda$}
    \end{subfigure}
    }
    \centerline{
    \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_alpha.eps}
        \caption{$\alpha$}
    \end{subfigure}
        \begin{subfigure}[b]{\subSize}
        \includegraphics[width=\textwidth]{../figures/compoundpoisson/CpEmAlgorithm_CompoundPoissonlambda1alpha1beta1_beta.eps}
        \caption{$\beta$}
    \end{subfigure}
    }
    \caption{The EM algorithm was used to estimate the parameters of a $\CPoisson(1,1,1)$ random variable using 1\,000 simulated samples. The graphs show the log-likelihood and the estimated parameters at each EM step for 10 different simulations. The dashed lines show the standard deviation using the Cram\'er-Rao lower bound.}
    \label{fig:compoundPoisson_convergence_1}
\end{figure}
